---
title: "Peer-graded Assignment: Statistical Inference Course Project"
author: "Kirill Setdekov"
date: '13 may 2019 г '
output:
  pdf_document:
    latex_engine: xelatex
---
# Overview

In this second part of the assignment, I will explore the tooth growth dataset and apply statistical inference tools learned from the course. There is statistically significant effect of dosage on tooth growth, but delivery method shows a difference only on one of the tests.
```{r libload, message=FALSE, warning=FALSE}
#load libraries silently
require(knitr)
require(ggplot2)
require(ggpubr)
require(gridExtra)
require(dplyr)
```

# Part 2. Basic inferential data analysis.
"Tooth" dataset consists of 60 observatiosn of measurements of tooth growth related cells in 60 guinea pigs. There were 3 dose levels of vitamin C (encoded as "dose") and 2 types of delivery methods, encoded as "supp", where VC is vitamin C and OJ is orange juice.

```{r tooth_exploratory, message=FALSE, warning=FALSE, fig.height = 3}
tooth <- datasets::ToothGrowth
summary(tooth)



p1 <-
    ggplot(tooth, aes(x = dose, y = len, color = supp)) +
    geom_point() + geom_smooth() +
    labs(title = "Length distribution by dose and \nsupp with trend lines")
p2 <-
    ggplot(tooth, aes(y = len, x = supp, fill = supp)) + geom_violin() +
    labs(title  = "Violin plot without dose \ndifferentiation")
grid.arrange(p1, p2, nrow = 1)
```

On the plots we can see that Vitamin C leads to higher variance in tooth growth in total.
If we look at mean length by supp and dose, we can see that it increases with dose and is higher for OJ than for VC for all doses, excep 2.0 mg.
```{r dosevssupp}
kable(tooth %>% group_by(supp, dose) %>% summarise(mean(len)))
```


## testing
### Tooth growth vs supp testing

```{r supp}
t.test(len~supp, paired = FALSE, var.equal = FALSE, data = tooth)

```
We conduct t-test for difference in means between group OJ and VC. We can see that the confidence interval includes zero and p-value > 0.05. We cannot reject the $H_0$ of no difference between these groups in terms of mean tooth growth.

However, if we  look at permutation testing and compare mean tooth growth for OJ and VC, we can reject $H_0$ at $\alpha = 0.05$.
```{r permutation}
toothshort <- tooth
y <- toothshort$len
group <- as.character(toothshort$supp)
testStat <- function(w, g)
    mean(w[g == "OJ"]) - mean(w[g == "VC"])
observedStat <- testStat(y, group)
permutations <-
    sapply(1:10000, function(i)
        testStat(y, sample(group)))
# difference in average count between b and c
observedStat
# share of more extreme
mean(permutations > observedStat)
# pvalue is very low - reject H0
```

### Tooth growth vs dose testing

```{r dose}
tooth2doses <- subset(tooth, dose %in% c(0.5,2))
t.test(len~dose, paired = FALSE, var.equal = FALSE, data = tooth2doses)
```
We conduct t-test for difference in means between group doses of 0.5 and 2 mg. We can see that p-value < 0.01. We can reject the $H_0$ of no difference between mean tooth growh for low and high dosages of Vitamin C.

## conclusions
Based on t-test and visual analysis of the data, we conclude that the amount of vitamin C consumed is linked to tooth growth, however the type of delivery in genereal does not lead to a statistically significant difference in tooth growth.

For these two tests I used the following assumptions:

* These are unpaired observations;
* Variance in unequal between tested groups.

Using perputation testing, we can say that at $\alpha = 0.05$, there is higher mean tooth growth for OJ, but this is not significant at a higher confidence level.