---
title: "Simulation and inferential data analysis"
author: "Kirill Setdekov"
date: '13 may 2019 г '
output:
  pdf_document:
    latex_engine: xelatex
---
# Part 1.A simulation exercise.
## Overview

In a few (2-3) sentences explain what is going to be reported on.

## Simulations: 
Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.

## Sample Mean versus Theoretical Mean: 
Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.

## Sample Variance versus Theoretical Variance: 
Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.

## Distribution: Via figures and text, explain how one can tell the distribution is approximately normal.



```{r}
hist(runif(1000))
```


```{r cars}
mns = NULL
for (i in 1 : 1000) mns = c(mns, mean(runif(40)))
hist(mns)
```

# Basic inferential data analysis.

```{r tooth_exploratory, message=FALSE, warning=FALSE, fig.height = 3}
tooth <- datasets::ToothGrowth

summary(tooth)

library(ggplot2)
library(gridExtra)
p1 <-
    ggplot(tooth, aes(x = dose, y = len, color = supp)) + geom_point() + geom_smooth() +
    labs(title = "Length distribution by dose and \nsupp with trend lines")
p2 <-
    ggplot(tooth, aes(y = len, x = supp, fill = supp)) + geom_violin() + labs(title  = "Violin plot without dose \ndifferentiation")
grid.arrange(p1, p2, nrow = 1)
```

## testing
Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)

## conclusions
State your conclusions and the assumptions needed for your conclusions.


# review criteria

* Did you show where the distribution is centered at and compare it to the theoretical center of the distribution?
* Did you show how variable it is and compare it to the theoretical variance of the distribution?
* Did you perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?
* Did the student perform some relevant confidence intervals and/or tests?
* Were the results of the tests and/or intervals interpreted in the context of the problem correctly?
* Did the student describe the assumptions needed for their conclusions?