# week 1
##  student T distribution
1 parameter, - degrees of freedom.
distribution of X average - mu/(S/sqrt(n))
with n-1 degrees of freedom
confidence interval interval - z+-t*S//sqrt(n)

For skewed distribution, the spirit of the t interval assumptions are violated.
  better to have logs or use different summary like the median

## T confidenct with pair measurements

```{r}
data(sleep)
head(sleep)
```

---
```{r}
g1 <- sleep$extra[1 : 10]; g2 <- sleep$extra[11 : 20]
difference <- g2 - g1
mn <- mean(difference); s <- sd(difference); n <- 10
mn + c(-1, 1) * qt(.975, n-1) * s / sqrt(n)
t.test(difference)$conf.int
```

## T Independent group T intervals

## Independent group $t$ confidence intervals

- Suppose that we want to compare the mean blood pressure between two groups in a randomized trial; those who received the treatment to those who received a placebo
- We cannot use the paired t test because the groups are independent and may have different sample sizes
- We now present methods for comparing independent groups

---

## Notation

- Let $X_1,\ldots,X_{n_x}$ be iid $N(\mu_x,\sigma^2)$
- Let $Y_1,\ldots,Y_{n_y}$ be iid $N(\mu_y, \sigma^2)$
- Let $\bar X$, $\bar Y$, $S_x$, $S_y$ be the means and standard deviations
- Using the fact that linear combinations of normals are again normal, we know that $\bar Y - \bar X$ is also normal with mean $\mu_y - \mu_x$ and variance $\sigma^2 (\frac{1}{n_x} + \frac{1}{n_y})$
- The pooled variance estimator $$S_p^2 = \{(n_x - 1) S_x^2 + (n_y - 1) S_y^2\}/(n_x + n_y - 2)$$ is a good estimator of $\sigma^2$

---

## Note

- The pooled estimator is a mixture of the group variances, placing greater weight on whichever has a larger sample size
- If the sample sizes are the same the pooled variance estimate is the average of the group variances
- The pooled estimator is unbiased
$$
    \begin{eqnarray*}
    E[S_p^2] & = & \frac{(n_x - 1) E[S_x^2] + (n_y - 1) E[S_y^2]}{n_x + n_y - 2}\\
            & = & \frac{(n_x - 1)\sigma^2 + (n_y - 1)\sigma^2}{n_x + n_y - 2}
    \end{eqnarray*}
$$
- The pooled variance  estimate is independent of $\bar Y - \bar X$ since $S_x$ is independent of $\bar X$ and $S_y$ is independent of $\bar Y$ and the groups are independent

---

## Result

- The sum of two independent Chi-squared random variables is Chi-squared with degrees of freedom equal to the sum of the degrees of freedom of the summands
- Therefore
$$
    \begin{eqnarray*}
      (n_x + n_y - 2) S_p^2 / \sigma^2 & = & (n_x - 1)S_x^2 /\sigma^2 + (n_y - 1)S_y^2/\sigma^2 \\ \\
      & = & \chi^2_{n_x - 1} + \chi^2_{n_y-1} \\ \\
      & = & \chi^2_{n_x + n_y - 2}
    \end{eqnarray*}
$$

---

## Putting this all together

- The statistic
$$
    \frac{\frac{\bar Y - \bar X - (\mu_y - \mu_x)}{\sigma \left(\frac{1}{n_x} + \frac{1}{n_y}\right)^{1/2}}}%
    {\sqrt{\frac{(n_x + n_y - 2) S_p^2}{(n_x + n_y - 2)\sigma^2}}}
    = \frac{\bar Y - \bar X - (\mu_y - \mu_x)}{S_p \left(\frac{1}{n_x} + \frac{1}{n_y}\right)^{1/2}}
$$
is a standard normal divided by the square root of an independent Chi-squared divided by its degrees of freedom 
- Therefore this statistic follows Gosset's $t$ distribution with $n_x + n_y - 2$ degrees of freedom
- Notice the form is (estimator - true value) / SE

---

## Confidence interval

- Therefore a $(1 - \alpha)\times 100\%$ confidence interval for $\mu_y - \mu_x$ is 
$$
    \bar Y - \bar X \pm t_{n_x + n_y - 2, 1 - \alpha/2}S_p\left(\frac{1}{n_x} + \frac{1}{n_y}\right)^{1/2}
$$
- Remember this interval is assuming a constant variance across the two groups
- If there is some doubt, assume a different variance per group, which we will discuss later


- Comparing SBP for 8 oral contraceptive users versus 21 controls
- $\bar X_{OC} = 132.86$ mmHg with $s_{OC} = 15.34$ mmHg
- $\bar X_{C} = 127.44$ mmHg with $s_{C} = 18.23$ mmHg
- Pooled variance estimate
```{r}
sp <- sqrt((7 * 15.34^2 + 20 * 18.23^2) / (8 + 21 - 2))
132.86 - 127.44 + c(-1, 1) * qt(.975, 27) * sp * (1 / 8 + 1 / 21)^.5
```

### lets mistreat sleep data as group
```{r}
data(sleep)
x1 <- sleep$extra[sleep$group == 1]
x2 <- sleep$extra[sleep$group == 2]
n1 <- length(x1)
n2 <- length(x2)
sp <- sqrt( ((n1 - 1) * sd(x1)^2 + (n2-1) * sd(x2)^2) / (n1 + n2-2))
md <- mean(x1) - mean(x2)
semd <- sp * sqrt(1 / n1 + 1/n2)
md + c(-1, 1) * qt(.975, n1 + n2 - 2) * semd
t.test(x1, x2, paired = FALSE, var.equal = TRUE)$conf
t.test(x1, x2, paired = TRUE)$conf
```


### chick weights

```{r}
require(party)
library(datasets)
require(reshape2)
require(ggplot2)

data("ChickWeight")
chick <- mob(data= ChickWeight, weight~Time|Diet, model = glinearModel, family=gaussian(link = "log"))
plot(chick)

wideCW <- dcast(ChickWeight, Diet+Chick~Time, value.var = "weight")
require(dplyr)
names(wideCW)[-(1:2)] <- paste("time",names(wideCW)[-(1:2)], sep = "")
wideCW <- wideCW %>% mutate(gain = time21-time0)

ggplot(data=wideCW, aes(x= factor(Diet), y = gain, fill=Diet))+geom_violin()

wideCW14 <- subset(wideCW, Diet %in% c(1,4))
rbind(t.test(gain~Diet, paired = FALSE, var.equal = TRUE, data = wideCW14)$conf,t.test(gain~Diet, paired = FALSE, var.equal = FALSE, data = wideCW14)$conf)
```

### Unequal variances.
when in doubt - use unequal variance calculation.
in R, t.test(..., var.equal = FALSE)

