---
title: "Statistic inference lecture"
author: "Kirill Setdekov"
date: "March 10, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
"Generating conclusions about a populations from a noisy sample"

*pick frequentist paradigm.*
[link to main repository](https://github.com/ksetdekov/courses/tree/master/06_StatisticalInference)

```{r libload, message=FALSE, include=FALSE}
#require(swirl)
#install_from_swirl("Statistical Inference")
```

## Probability

$$P(A\cup B)=P(A)+P(B)$$

$$ P(A \cup B)=P(A)+P(B)-P(A \cap B)$$

## probability mass function and density function

* PMF
    * must be larger or equal to 0
    * the sum of possible values sum up to 1
* PDF 
    * must be larger or equal to 0
    * the sum of possible values sum up to 1. Area under PDFs correspond to probabilities for that random variables

    
```{r}
x <- c(-0.5,0,1,1,1.5)
y <- c(0,0,2,0,0)
plot(x,y, lwd=3, frame = FALSE, type="l")


```

## CDF and survival function

- The **cumulative distribution function** (CDF) of a random variable $X$ is defined as the function 
$$
F(x) = P(X \leq x)
$$
- This definition applies regardless of whether $X$ is discrete or continuous.
- The **survival function** of a random variable $X$ is defined as
$$
S(x) = P(X > x)
$$
- Notice that $S(x) = 1 - F(x)$
- For continuous random variables, the PDF is the derivative of the CDF


$$
S(x) = 1 - x^2
$$

```{r}
pbeta(c(0.4, 0.5, 0.6), 2, 1)
```

## Quantiles

You've heard of sample quantiles. If you were the 95th percentile on an exam, you know
that 95% of people scored worse than you and 5% scored better. 
These are sample quantities. Here we define their population analogs.


---
## Definition

The  $\alpha^{th}$ **quantile** of a distribution with distribution function $F$ is the point $x_\alpha$ so that
$$
F(x_\alpha) = \alpha
$$
- A **percentile** is simply a quantile with $\alpha$ expressed as a percent
- The **median** is the $50^{th}$ percentile

---

## Example
What is the median of the distribution that we were working with before?
- We want to solve $0.5 = F(x) = x^2$
- Resulting in the solution 
```{r, echo = TRUE} 
sqrt(0.5)
``` 
- Therefore, about `r sqrt(0.5)` of calls being answered on a random day is the median.

---
## Example continued
R can approximate quantiles for you for common distributions

```{r}
qbeta(0.5, 2, 1) #quantiles
```


## summary
- We're referring to are **population quantities**. Therefore, the median being
  discussed is the **population median**.
- A probability model connects the data to the population using assumptions.
- Therefore the median we're discussing is the **estimand**, the sample median will be the **estimator**