---
title: "Which transmission type gets better MPG and by how much?"
author: "Kirill Setdekov"
date: "September 11, 2019"
output:
  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	cache = TRUE,
	echo = TRUE
)
```
```{r load, message=FALSE, warning=FALSE, include=FALSE}
require(datasets)
require(ggplot2)
require(dplyr)
require(gridExtra)
data(mtcars)
```
```{r preparation}
mtcars <- mtcars %>%
    mutate(
        wt = wt * 0.45359237,
        am = factor(am, levels = c(0, 1), labels = c("AT", "MT")),
        cylsize = disp / cyl
    ) # convert weight to metric t and create cylinder size variable
```
## Summary

According to the analysis, manual transmission vehicles have 7.245 better MPG compared to automatic transmission ones. However, adjusting the MPG by other variables, transmission type does not play a statistically significant role in MPG, having other factors constant. This suggests that not manual transmission leads to better MPG, but lighter cars with less cylinders and smaller cylinders.



```{r expl, echo=FALSE, fig.height=3, fig.width= 6, message=FALSE, warning=FALSE}
# exploratory analysis
p1 <-
ggplot(mtcars, aes(y = mpg, x = am, fill = am)) + geom_violin() +
labs(title = "Violin plot of MPG \nby transmission type")+theme_bw()

p2 <-
ggplot(mtcars, aes(x = wt, y = mpg, color = am)) +
geom_point() + geom_smooth() +
labs(title = "MPG distribution by wt and \ntransmission type")+
    theme_bw()

grid.arrange(p1, p2, nrow = 1)
```

We can see that on average manual cars get better MPG, however if wee look at MPG vs weight for both transmission types, we see that:
1. on general MT cars are lighter;
2. there is a significant overlap of confidence intervals for the region of 1 to 1,6 t weights.

## Regression analysis
I performed several sets of regression.

### Basis mpg vs transmission type

Results are significant, residuals look normal and no outliers.

```{r regressionamonly}
mtam <- lm(mpg~am, data = mtcars)
summary(mtam)
```

### All original factors included

None of the coefficients are significant, but we find, that cyl, displ, wt are individually contributing to a good model (also, cylsize, not shown).

```{r regressions}
mt <- lm(mpg~. -cylsize, data = mtcars)
summary(mt)
anova(mt)
```


### Best model search

Model mt3 is the best found one that includes and does not loose according to anova compared to a model with all factors. There are also no significant issues with residuals, however they fail the normality test. However, exluding transmission type `am` in a model mt3 is also possible, which again suggests that given other factors, transmission type does not affect MPG.
```{r bestmodel}
mt3 <- lm(mpg~disp+cyl+wt+cylsize+am, data = mtcars)
summary(mt3)
anova(mt3,mt)
shapiro.test(mt3$residuals)
par(mfrow = c(2, 2))
plot(mt3)

mt4 <- lm(mpg~disp+cyl+wt+cylsize, data = mtcars)
anova(mt3,mt4)
```


## Numeric results and comparisons

```{r compare, echo=FALSE, fig.height=4, fig.width=7, dpi=300}
mancoef <- c(mtam$coefficients[2],mt$coefficients[9],mt3$coefficients[6])
confmanlower <- c(confint(mtam)[2,1],confint(mt)[9,1],confint(mt3)[6,1])
confmanupper <- c(confint(mtam)[2,2],confint(mt)[9,2],confint(mt3)[6,2])
estimate <- c("Transmission only","Adjusted for all factors","Adjusted by best model")
coefdf <- data.frame(estimate,mancoef,confmanlower, confmanupper)
coefdf$estimate <- relevel(coefdf$estimate,"Transmission only")


p <- ggplot(coefdf, aes(x=estimate, y=mancoef, fill=estimate)) + 
  geom_bar(stat="identity", color="black") +
  geom_errorbar(aes(ymin=confmanlower, ymax=confmanupper), width=.3,
                 position=position_dodge(.9))
p+labs(title="Comparison of manual transmission impact on MPG \nwith confidence intervals", x="Model used", y = "Change in MPG for manual compared to AT")+
   theme_bw() 
```
By itsetlf manual transmission yields ```r formatC(mtam$coefficients[2]) ``` higher MPG compared to an automatic transmission

Adjusted for all factors - manual transmissions in a car gives it `r formatC(mt$coefficients[9])` better MPG compared to an automatic transmission.
However, this result is statistically insignificant and the best predictors of MPG are cylinder count, displacement, cylinder size and weight.
Adjusting for the factors in the best model, manual transmission provides `r formatC(mt3$coefficients[6])` lower MPG compared to an automatic transmission. 

In both cases effect of manual vs auto transmission, given all other factors equal is statistically indistinguishible from 0, while other factors remain significant.
