---
title: "NOAA storm event impact analysis"
author: "Kirill Setdekov"
date: "February 28, 2019"
output: html_document
---

## Synopsis
In the following article I conduct analysis on the NOAA data to determine the most damagind natural events for the  USA in terms of monetary impact as well as injuries and death.

I proceed to load original data, process it and determine 10 top most impactful event types in monetary tems (both for property and crop damage). 

Next the events' impact on injuries and deaths are calculated nationwide again to show 10 most ipmactful disaster types.

To get a broader picture, i calculate the most deadly events for each state and plot them on the map of the USA.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing 
### Load libraries
```{r liblary_load}
library(ggplot2)
library(dplyr)
library(reshape2)
library(usmap)
library(formatR)
```


### Data loading
We cache the calculation and load data from the link and directrly read it into R.
```{r data_load, cache=TRUE}
link="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(link, "data/noaa.bz2")
noaa <- read.csv(bzfile("data/noaa.bz2"))
```
To plot data we need to summarize on an event-type level.
### material damage
I use `dplyr` library to calculate crop and property damage and their sum for each of the event types for the whole dataset. After that, events are ordered by total damage and data is further transformed to be plotted.

### death and injury
For population health effects I chose injury and death data. Similarly it is summarized by event type in terms of damages. Next events are ordered by humal losses to emphasize more deadly ones and data is further transformed to be plotted.

### map data preparation
Infrormation about human lossed is summarised by event and state, then then for each state only a single event with the highes toll is selected. This information is matched by 2-letter state codes with an existing dataset of 51 states that is easily plotted on the map to indicated the mose deadly event for each of the states.

```{r data_preparation, tidy=TRUE}
## material damage #####
getExp <- function(x) {
    ifelse (x %in% c("h", "H"), return(2), {
    ifelse (x %in% c("k", "K"), return(3), {
    ifelse (x %in% c("m", "M"), return(6), {
    ifelse (x %in% c("b", "B"), return(9), {
    ifelse (!is.na(as.numeric(x)), return(as.numeric(x)),return(0))
    
    })
    })
    })
    })
    }
    
noaa<- noaa%>% mutate(PROPDMG=PROPDMG*(10^(getExp(PROPDMGEXP))))
noaa<- noaa%>% mutate(CROPDMG=CROPDMG*(10^(getExp(CROPDMGEXP))))



damages <- noaa %>% group_by(EVTYPE) %>% summarise(property=sum(PROPDMG), crop=sum(CROPDMG), total=sum(PROPDMG+CROPDMG))
damages <- damages[order(-damages$total),]
damages$EVTYPE <- factor(damages$EVTYPE, levels = damages$EVTYPE[order(-damages$total)])
damages_long <- melt(damages[1:10,c(1,2:3)], id = "EVTYPE")
damages_long$EVTYPE <- droplevels(damages_long$EVTYPE)

## injuries and death data calculation #####                                
people <- noaa %>% group_by(EVTYPE) %>% summarise(injuries=sum(INJURIES), fatalities=sum(FATALITIES), total=sum(INJURIES+FATALITIES))
people <- people[order(-people$total),]
people$EVTYPE <- factor(people$EVTYPE, levels = people$EVTYPE[order(-people$fatalities)])
people_long <- melt(people[1:10,c(1,2:3)], id = "EVTYPE")
people_long$EVTYPE <- droplevels(people_long$EVTYPE)

## calculation of data for the map  #####                                
temp <-
noaa  %>% group_by(STATE, EVTYPE) %>% summarise(FATALITIES = sum(FATALITIES)) %>%
    filter(FATALITIES == max(FATALITIES)) %>% top_n(n =1, FATALITIES) %>% slice(1)
## using an alredy present dataframe with map and states from the library
randompop <- statepop
## allocate the most important event here
randompop <- left_join(randompop, temp, by = c("abbr" = "STATE"))
randompop$EVTYPE <- as.character(randompop$EVTYPE)

bystates <-
(randompop %>% group_by(EVTYPE) %>% summarise(sum = sum(FATALITIES)))
randompop$EVTYPE <-
factor(randompop$EVTYPE, levels = bystates$EVTYPE[order(-bystates$sum)])
```
```{r barplots}
## demage
ggplot(data = damages_long, aes(x = EVTYPE, y = value ,
                                fill = variable)) +
    geom_bar(stat = "identity") +
    ylab("Total property damage") +
    labs(fill = "Total damage breakdown",
         title = "10 disaster types sorted by total damage for all states") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
## death and injury
ggplot(data = people_long, aes(x = EVTYPE, y = value ,
                               fill = variable)) +
    geom_bar(stat = "identity") +
    ylab("People count") + 
    labs(fill = "Total death+injury", title = "10 disaster types sorted by fatalities for all states") +
    theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


``` {r usa_plot, echo = FALSE, fig.width = 12, fig.height = 9, out.width = "1200", out.height = "900"}
# plot
plot_usmap(data = randompop, values = "EVTYPE") + 
  labs(title = "US States", subtitle = "most deadly event for each state") + 
  theme(panel.background = element_rect(colour = "black", fill = "lightblue"))


```

## Results 
Based on the performed calculations and available data, most deadly events are:

1. Tornados
2. Excessive heat
3. Flash floods

Based on economic efect only, the events with the highes economic consequences are these three:

1. Hail 
2. Tornados
3. Flash floods

Looking at the state distribution of the most deadly event, we can find possible explaination based on local climat and geographic features of the states. Examples include tornados in the eastern US and excessive heat in California. Some states reveal non-obvious results, for exapmle, lighning causing the most deaths overall.
