My firs R markdown file
==============
This is my firs R marksown file
```{r mean of random}
set.seed(1)
x <- rnorm(100)
mean(x)
```

Here we're going to load some data
```{r}
library(datasets)
data("airquality")
summary(airquality)
```

Lets's firs make a plot pairwise

```{r pairs, echo=FALSE}
pairs(airquality)
```

Here's a regression model of ozone on wind, solar raditaion and temp

```{r}
library(stats)
fit <- lm(Ozone~Wind+Solar.R+Temp, data = airquality)
summary(fit)
```

Here's an unordered list:

* first

* second

## can have inline text computations

```{r computetime, echo=FALSE}
time <- format(Sys.time(), "%a %b %d %X %Y")
rand <- rnorm(1)
```
The current time is `r time`. Random number of the day is `r rand`.

## plots

let's gen data
```{r simdata, echo=T}
x <- rnorm(100); y <- x+rnorm(100, sd=0.5)
```
Here's a scatterplot
```{r scatterplot, fig.height=4}
library(ggplot2)
qplot(x,y, main="My Sim data")
```

## tables

```{r fitmodel}
library(datasets)
data(airquality)
fit <- lm(Ozone~Wind+Solar.R+Temp, data = airquality)
```
Here's a table of regression coefficients
```{r showtable, results='asis'}
library(xtable)
xt <- xtable(summary(fit))
print(xt, type="html")
```